Mechanization of Separation in Generic Extensions

### 2019-01-07 LICS/conclusion_m.tex
otra vuelta más a forcing_thms
kaiser & barras en concl_m. absolutez en powerset

We cannot overstate the importance of following the sharp and detailed 
presentation of forcing given by
\citet{kunen2011set}. In fact, it helped us to delineate the
\emph{thematic} aspects of our formalization; i.e.~the handling of all
the theoretical concepts and results in the subject and it informed
the structure of locales organizing our development. This had an
impact, in particular, in the formal statement of the Fundamental
Theorems. We consider that the writing of the \isatt{forcing\_thms}
locale, though only taking a few lines of code, is the second most
important achievement of this work, since there is no obvious
reference from which to translate this directly.

@@

This is an appropriate point to insist that the internal/external
dichotomy has been a powerful agent in the shaping of our project.
This tension was also pondered by Paulson in his formalization of
G\"odel's constructible universe \cite{paulson_2003}; after choosing a
shallow embedding of $\ZF$, every argument proved by induction on
formulas (or functions defined by recursion) should be done using
internalized formulas. Working on top of Paulson's library, we
prototyped the formula-transformer $\forceisa$, which is defined for
internalized formulas, and this affects indirectly the proof of the
Separation Axiom (despite the latter is not by induction). The proof
of Replacement also calls for internalized formulas, because one needs
a general version of the Reflection Principle (since the formula
$\phi$ involved depends on the specific instance of Replacement being
proved).

>>>>

We cannot overstate the importance of following the sharp and detailed 
presentation of forcing given by
\citet{kunen2011set}. In fact, it helped us to delineate the
\emph{thematic} aspects of our formalization; i.e.~the handling of all
the theoretical concepts and results in the subject and it informed
the structure of locales organizing our development. This had an
impact, in particular, in the formal statement of the Fundamental
Theorems. We consider that the writing of the \isatt{forcing\_thms}
locale, though only taking a few lines of code, is the second most
important achievement of this work, since there is no obvious
reference from which to translate this directly. The accomplishment
of the formalizations of Separation and Powerset are, in a sense,
certificates that the locale of the Fundamental Theorems was set
correctly.

@@

This is an appropriate point to insist that the internal/external
dichotomy has been a powerful agent in the shaping of our project.
This tension was also pondered by Paulson in his formalization of
G\"odel's constructible universe \cite{paulson_2003}; after choosing a
shallow embedding of $\ZF$, every argument proved by induction on
formulas (or functions defined by recursion) should be done using
internalized formulas. Working on top of Paulson's library, we
prototyped the formula-transformer $\forceisa$, which is defined for
internalized formulas, and this affects indirectly the proof of the
Separation Axiom (despite the latter is not by induction). The proof
of Replacement also calls for internalized formulas, because one needs
a general version of the Reflection Principle (since the formula
$\phi$ involved depends on the specific instance of Replacement being
proved). Other formalizations, v.g.\ that of Barras
\cite{barras-habilit} and Kaiser \cite{kaiser-set-theory-coq}, take
advantage of a better articulation between the metatheory (Coq in both
cases) and the set theory. An alternative road would be to redevelop
absoluteness 
results in a more structured metatheory that already includes a
recursively defined type of first order formulas. Needless to say, this
comprises an extensive re-engineering.
